---
title: "Categorical Data"
author: "Syed Arham Naqvi"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
---

## About
This notebook will utilize visualization tools including bar graphs or histograms, Venn diagrams and two-way tables to analyze patterns or relationships in categorical data.


### Initializing Environment and Loading Libraries
```{r}
graphics.off()
rm(list = ls())
library(patchwork)
library(tidyverse)
```

## Dog Breed Histogram
```{r, fig.height=3, fig.width=5}
d <- tibble(
  color = c("Blue", "Green", "Red", "Yellow", "Pink"),
  n = c(60, 105, 45, 90, 90)
)

ggplot(d, aes(x=color, y=n))+
  geom_col(width = 0.7, fill = d$color) +
  scale_y_continuous(limits = c(0, max(d$n)+10),
                     breaks = seq(from=0, to=max(d$n)+10, by = 10),
                     expand = c(0,0))+
  theme_bw() +
  labs(x="Eye Color",y="Number of Students", title = "Number of Studnets per Eye Color")
```

## Test Grades Bar Chart
```{r, fig.height=2, fig.width=4}

get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

d <- tibble(
  name = c("Ishaan", "Emily", "Daniel", "Jessica", "William"),
  midterm = c(90, 90, 60, 100, 80),
  final = c(100, 100, 100, 75, 80)
  )
d_long <- d |> 
  pivot_longer(
    cols = c(midterm, final),
    values_to = "grade",
    names_to = "test_type"
  ) |> 
  mutate(test_type = factor(test_type, levels = c("midterm", "final")))

ggplot(d_long, aes(x = name, y = grade, fill = test_type))+
  geom_col(width = 0.7, position = position_dodge(width = 0.7))

summary <- d |> 
  summarise(final_median = median(final),
            midterm_midrange = (max(midterm) + min(midterm))/2,
            final_avg = mean(final),
            final_mode = get_mode(final),
            midterm_range = max(midterm)-min(midterm))
summary <- summary |> 
  pivot_longer(cols = names(summary),
               names_to = "stat",
               values_to = "value")
summary
ggplot(summary, aes(y=stat, x=value)) +
  geom_col(color = "black", alpha=0) +
  theme_bw()

```

## Monthly Sales Percentage Pie Chart
```{r, fig.height=5, fig.width=5}

library(ggrepel)

months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
months <- factor(months, levels = months)

s <- tibble(
  month = months,
  percent_sales = c(0.18, 0.1, 0.07, 0.05, 0.05, 0.03, 0.03, 0.09, 0.07, 0.08, 0.13, 0.12)
) |> 
  arrange(desc(month)) |> 
  mutate(cumsum = cumsum(percent_sales),
         pos = cumsum - percent_sales/2,
         label = scales::percent(percent_sales))
s |> 
  ggplot(aes(x="", y=percent_sales, fill = month)) +
  geom_col(width=1, color="black", linewidth = 0.2) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Set3") +
  geom_label_repel(aes(y = pos, label = label),
                   size = 2.5,
                   nudge_x = 0.7,
                   show.legend = FALSE) +
  theme_minimal() +
  theme(axis.line = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

## Creating Frequency Table Using Given Information

Kirti knows the following information from a study on cold medicine that included 60 participants:

- 30 participants in total received cold medicine.
- 26 participants in total had a cold that lasted longer than 7 days.
- 14 participants received cold medicine but had a cold that lasted longer than 7 days.

```{r}

a11 <- 14
a21 <- 30 - 14
a12 <- 26 - 14
a22 <- 60 - 26 - a21

two_way_matrix <- matrix(c(a11, a12, a21, a22), nrow = 2, byrow = TRUE)
two_way_table <- as.table(two_way_matrix)
dimnames(two_way_table) <- list('Duration of Cold' = c('longer than 7 days', 'shorter than 7 days'),
                                'Medicine' = c('Received', 'Not Received'))
print(two_way_table)

```

## Conditional Distribution Stacked-Bar Plot
```{r}
d <- tibble(
  PetSource = c("Store or breeder", "Stray", "Shelter", "Family or friends"),
  Dogs = c(65, 11, 65, 35),
  Cats = c(5, 30, 49, 31),
  Total = c(70, 41, 114, 66)
)

d

d <- d |> 
  pivot_longer(c('Dogs', 'Cats'), names_to = 'pet', values_to = 'count') |> 
  relocate(Total, .after=count) |> 
  mutate(percent_of_total = count/Total) |> 
  mutate(Total = NULL)

d |> 
  ggplot(aes(x=PetSource, y=percent_of_total, fill=pet)) +
  geom_col(color = "black", linewidth=0.2) +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent) +
  theme_bw()

```


```{r}
round((2*85+84+83+81)/5, digits = 2)

```








